Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-07-07 manuscript/manuscript.tex
create pdf of ms after review by Rich

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such comparisons between means have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development as well as our understanding of the structure of cognition and brain function. Indeed, the view that mean values are truth and variations around the mean are error is deeply ingrained in experimental psychology (Bolger, Zee, Rossignac-Milon, \& Hassin, 2019). However, differences in mean RT conceal important pieces of information, such as when an experimental effect starts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events (Panis, 2020; Panis, Moran, Wolkersdorfer, \& Schmidt, 2020; Panis \& Schmidt, 2016, 2022; Panis, Torfs, Gillebert, Wagemans, \& Humphreys, 2017; Panis \& Wagemans, 2009; Wolkersdorfer, Panis, \& Schmidt, 2020). Such absolute timing information is useful not only for the interpretation of experimental effects under investigation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

As a simple illustration, Figure 1 summarises simulated single-subject data (200 trials per condition) that shows how comparing means between two conditions can conceal the shapes of the underlying RT and accuracy distributions. Indeed, compared to the aggregation of data across trials (Figure 1A), a distributional approach offers the possibility to reveal the time course of psychological states (Figure 1B). For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higher in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (\textgreater500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1 (see also Panis \& Schmidt, 2016).

@@

Why does this matter for research in psychology? For many psychological questions, the estimation of such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes. Because EHA adds a relatively under-used but ever-present dimension -- the passage of time -- to the theory building toolkit, it provides one possible answer to the recent call for a temporal science of behavior (Abney, Fausey, Suarez-Rivera, \& Tamis-LeMonda, 2025).

@@

Our ultimate aim in this paper is twofold: first, we want to convince readers of the many benefits of using EHA when dealing with psychological RT data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a (single event) discrete-time EHA on RT data, as well as a complementary discrete-time speed-accuracy tradeoff (SAT) analysis on timed accuracy data in case of choice RT data (Figure 1B).

Even though EHA is a widely used statistical tool and there already exist many excellent reviews (Allison, 1982; Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; Mills, 2011; Singer \& Willett, 2003; Teachman, 1983) and tutorials (Allison, 2010; Elmer, Van Duijn, Ram, \& Bringmann, 2023; Landes, Engelhardt, \& Pelletier, 2020; Lougheed, Benson, Cole, \& Ram, 2019; Stoolmiller, 2015; Stoolmiller \& Snyder, 2006), we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain \textbf{how} to do such analyses in the context of experimental psychology, rather than repeat in any detail \textbf{why} you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations (Allison, 2010). In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics (Allison, 2010; Gelman, Hill, \& Vehtari, 2020; Mills, 2011; Singer \& Willett, 2003; Winter, 2019). We supply a few regression equations in section E of the Supplemental Material.

@@

There are different types of modeling approaches in EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper.
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable.
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well (Bloxom, 1984; Luce, 1991; Van Zandt, 2000).
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than \textasciitilde200 trials per condition per participant (Panis, Schmidt, et al., 2020).
Moreover, as indicated by Allison (2010), learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

To apply discrete-time EHA, one divides the within-trial time in discrete, contiguous time bins indexed by t (e.g., t = 1 to 10 time bins; Figure 1B).
Then let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular trial (i.e., repeated measure).
For example, a response in one trial might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).
One then calculates the sample-based estimate of the discrete-time hazard function of event occurrence for each experimental condition (Figure 1B top).
The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in
bin t, given that the event does not occur in previous bins.
In other words, it reflects the instantaneous risk that the event occurs in the current bin t, given that it has not yet
occurred in the past, i.e., in one of the prior bins (t-1, t-2, \ldots, 1).

In the context of experimental psychology, it is often (but not always), the case that responses can be classified as correct or incorrect.
In those cases, one can also calculate the conditional accuracy function (Figure 1B bottom).
The conditional accuracy function gives you for each time bin the conditional probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).
The conditional accuracy function is also known as the micro-level speed-accuracy tradeoff (SAT) function.
We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT. The definitions of these and other discrete-time functions are given in section B of the Supplemental Material.

@@

Second, compared to more conventional analytical approaches, EHA uses more of the data because it deals with missing data differently.
It is conventional with RT data to either (a) use a response deadline and discard all trials without a response, or (b) wait in each trial until a response occurs and then apply data trimming techniques, i.e., discarding too short or too long RTs (and perhaps also erroneous responses) before calculating a mean RT (Berger \& Kiefer, 2021).
Discarding data can introduce biases, however.
Rather than treat non-responses as missing data, EHA treats such trials as \emph{right-censored} observations on the variable RT, because all we know is that RT is greater than some value.
Right-censoring is a type of missing data problem and a nearly universal feature of survival data including RT data.
For example, if the censoring time was 1 second, then some trials result in observed event times (those with a RT below 1 second), while the other trials result in response times that are right-censored at 1 second.
The fact that EHA can deal with right-censoring, therefore, presents a analytical strength of the approach compared to many common approaches in experimental psychology (ANOVA, linear regression, delta plots).

@@

Third, the width of each time bin will need to be determined.
For instance, in Figure 1B we chose 100 ms in an arbitrary manner.
In reality, however, bin width will need to be set by considering a number of factors simultaneously.
The optimal bin width will depend on (a) the length of the observation period in each trial, (b) the rarity of event occurrence, (c) the number of repeated measures (or trials) per condition per participant, and (d) the shape of the hazard function.
Finding an appropriate bin width in a given user case before fitting models will require testing a number of options, when calculating and plotting the descriptive statistics (see section 3.1). The goal is to find the smallest bin width that is supported by the amount of data available. Based on our experience, a bin width of 50 ms is a good starting value when the number of repeated measures is 100 or less. Too small bin widths will result in erratic hazard functions as many bins will have no events, and thus hazard estimates of zero. Interestingly, the time bins do not need to have the same width. For example, Panis (2020) used larger bins towards the end of the observation period, as fewer events occurred there.

@@

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.5.1; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.13.0; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.9.0.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.2; Wickham, 2016), \emph{lme4} (Version 1.1.37; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.4; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.3; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.168; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.4; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.14; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{rstan} (Version 2.32.7; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.3.0; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.5; Barth, 2023) for all reported analyses.

@@

Hazard and conditional accuracy models can quickly become very complex when adding more than one time scale, due to the many possible higher-order interactions.
For example, some of the models discussed in Tutorial 2a, which we did not focus on in the main text, contain two time scales as covariates: the passage of time on the within-trial time scale, and the passage of time on the across-trial (or within-experiment) time scale. However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises a number of sessions, then four time scales can be defined (within-trial, within-block, within-session, and within-experiment).
From a theoretical perspective, adding more than one time scale -- and their interactions -- can be important to capture plasticity and other learning effects that may play out on such longer time scales, and that are probably present in each experiment in general (Schöner \& Spencer, 2016).
From a practical perspective, therefore, some choices need to be made to balance the amount of data that is being collected per participant, condition and across the varying timescales. As one example, if there are several times cales of relevance, then it might be prudent for interpretational purposes to limit the number of experimental predictor variables (conditions).
This is of course where planning and data simulation efforts would be important to provide a guide to experimental design choices (see Tutorial 4 and section 2.3).

>>>>

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such comparisons between means have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development as well as our understanding of the structure of cognition and brain function. Indeed, the view that mean values represent truth and variations around the mean are error is deeply ingrained in experimental psychology (Bolger, Zee, Rossignac-Milon, \& Hassin, 2019). However, differences in mean RT conceal important pieces of information, such as when an experimental effect starts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events (Panis, 2020; Panis, Moran, Wolkersdorfer, \& Schmidt, 2020; Panis \& Schmidt, 2016, 2022; Panis, Torfs, Gillebert, Wagemans, \& Humphreys, 2017; Panis \& Wagemans, 2009; Wolkersdorfer, Panis, \& Schmidt, 2020). Such absolute timing information is useful not only for the interpretation of experimental effects under investigation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

As a simple illustration, Figure 1 summarises simulated data for one subject that shows how comparing means between two conditions can conceal the shapes of the underlying RT and accuracy distributions. Indeed, compared to the aggregation of data across trials (Figure 1A), a distributional approach offers the possibility to reveal the time course of psychological states (Figure 1B). For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higher in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (\textgreater500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1. Importantly from a face-validity perspective, this pattern of simulated data can be seen in the experimental psychology literature (Panis \& Schmidt, 2016).

@@

Why does this matter for research in psychology? For many psychological questions, the estimation of such ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes. Because EHA adds a relatively under-used but ever-present dimension -- the passage of time -- to the theory building toolkit, it provides one possible response to the recent call for a temporal science of behavior (Abney, Fausey, Suarez-Rivera, \& Tamis-LeMonda, 2025).

@@

Our ultimate aim in this paper is twofold. First, we want to convince readers of the many benefits of using EHA when dealing with psychological RT data. Second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a (single event) discrete-time EHA on RT data, as well as a complementary discrete-time speed-accuracy tradeoff (SAT) analysis on timed accuracy data in case of choice RT data (Figure 1B).

Even though EHA is a widely used statistical tool and there already exist many excellent reviews (Allison, 1982; Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; Mills, 2011; Singer \& Willett, 2003; Teachman, 1983) and tutorials (Allison, 2010; Elmer, Van Duijn, Ram, \& Bringmann, 2023; Landes, Engelhardt, \& Pelletier, 2020; Lougheed, Benson, Cole, \& Ram, 2019; Stoolmiller, 2015; Stoolmiller \& Snyder, 2006), we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps.

Set within this context, our overall aim is to introduce a set of tutorials, which explain \textbf{how} to do such analyses in the context of experimental psychology, rather than repeat in any detail \textbf{why} you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations (Allison, 2010). In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics (Allison, 2010; Gelman, Hill, \& Vehtari, 2020; Mills, 2011; Singer \& Willett, 2003; Winter, 2019). We also supply relevant regression equations in section E of the Supplemental Material.

@@

There are different types of modeling approaches in EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper.
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable.
However, continuous forms of EHA require much more data to reliably estimate the continuous-time hazard (rate) function (Bloxom, 1984; Luce, 1991; Van Zandt, 2000).
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than \textasciitilde200 trials per condition per participant (Panis, Schmidt, et al., 2020).
Moreover, as indicated by Allison (2010), learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

To apply discrete-time EHA, one divides the within-trial time in discrete, contiguous time bins indexed by t (e.g., t = 1 to 10 time bins; Figure 1B).
Then let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular trial (i.e., repeated measure).
For example, a response in one trial might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).
One then calculates the sample-based estimate of the discrete-time hazard function of event occurrence for each experimental condition (Figure 1B upper panel).
The discrete-time hazard function gives you, for each time bin, the conditional probability that the event occurs (sometime) in
bin t, given that the event does not occur in previous bins.
In other words, it reflects the instantaneous risk that the event occurs in the current bin t, given that it has not yet
occurred in the past, i.e., in one of the prior bins (t-1, t-2, \ldots, 1).

In the context of experimental psychology, it is often (but not always), the case that responses can be classified as correct or incorrect.
In those cases, one can also calculate the conditional accuracy function (Figure 1B lower panel).
The conditional accuracy function gives you for each time bin the conditional probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).
The conditional accuracy function is also known as the micro-level speed-accuracy tradeoff (SAT) function.
We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT. The definitions of these and other discrete-time functions are given in section B of the Supplemental Material.

@@

Second, compared to more conventional analytical approaches, EHA uses more of the data because it deals with missing data differently.
It is conventional with RT data to either (a) use a response deadline and discard all trials without a response, or (b) wait in each trial until a response occurs and then apply data trimming techniques, i.e., discarding too short or too long RTs (and perhaps also erroneous responses) before calculating a mean RT (Berger \& Kiefer, 2021).
Discarding data can introduce biases, however.
Rather than treat non-responses as missing data, EHA treats such trials as \emph{right-censored} observations on the variable RT, because all we know is that RT is greater than some value.
Right-censoring is a type of missing data problem and a nearly universal feature of survival data including RT data.
For example, if the censoring time was 1 second, then some trials result in observed event times (those with a RT below 1 second), while the other trials result in response times that are right-censored at 1 second.
The fact that EHA can deal with right-censoring, therefore, presents a analytical strength of the approach compared to many common approaches in experimental psychology (e.g., ANOVA, linear regression, delta plots).

@@

Third, the width of each time bin will need to be determined.
For instance, in Figure 1B we chose 100 ms in an arbitrary manner.
In reality, however, bin width will need to be set by considering a number of factors simultaneously.
The optimal bin width will depend on (a) the length of the observation period in each trial, (b) the rarity of event occurrence, (c) the number of repeated measures (or trials) per condition per participant, and (d) the shape of the hazard function.
Finding an appropriate bin width in a given user case before fitting models will require testing a number of options, when calculating and plotting the descriptive statistics (see section 3.1). The goal is to find the smallest bin width that is supported by the amount of data available. Based on our experience, a bin width of 50 ms is a good starting value when the number of repeated measures is 100 or less. Overly small bin widths will result in erratic hazard functions as many bins will have no events, and thus hazard estimates of zero. Of note, however, is that time bins do not need to have the same width. For example, Panis (2020) used larger bins towards the end of the observation period, as fewer events occurred there.

@@

The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2020), Heiss (2021), Kurz (2023a), and Kurz (2023b). We used R (Version 4.5.1; R Core Team, 2024) and the R-packages \emph{bayesplot} (Version 1.13.0; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.22.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.9.0.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.2; Wickham, 2016), \emph{lme4} (Version 1.1.37; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.4; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.3; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.168; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.3; Aust \& Barth, 2024), \emph{patchwork} (Version 1.3.0; Pedersen, 2024), \emph{purrr} (Version 1.0.4; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.14; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{rstan} (Version 2.32.7; Stan Development Team, 2024), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{StanHeaders} (Version 2.32.10; Stan Development Team, 2020), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.3.0; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.7; Kay, 2024), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019) and \emph{tinylabels} (Version 0.2.5; Barth, 2023) for all reported analyses.

@@

Hazard and conditional accuracy models can quickly become very complex when adding more than one time scale, due to the many possible higher-order interactions.
For example, some of the models discussed in Tutorial 2a, which we did not focus on in the main text, contain two time scales as covariates: the passage of time on the within-trial time scale, and the passage of time on the across-trial (or within-experiment) time scale. However, when trials are presented in blocks, and blocks of trials within sessions, and when the experiment comprises a number of sessions, then four time scales can be defined (within-trial, within-block, within-session, and within-experiment).
From a theoretical perspective, adding more than one time scale -- and their interactions -- can be important to capture plasticity and other learning effects that may play out on such longer time scales, and that are probably present in each experiment in general (Schöner \& Spencer, 2016).
From a practical perspective, therefore, some choices need to be made to balance the amount of data that is being collected per participant, condition and across the varying timescales. As one example, if there are several timescales of relevance, then it might be prudent for interpretational purposes to limit the number of experimental predictor variables (conditions).
This is of course where planning and data simulation efforts would be important to provide a guide to experimental design choices (see Tutorial 4 and section 2.3).
